---
title: "Microbenchmark: Architectural Justification"
subtitle: "Performance analysis of list-of-data.frames design for Chromatograms"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## System Information

```{r}
#| label: system-info-early
#| code-fold: true
cat("CPU Cores:", parallel::detectCores(), "\n")
cat("R Version:", R.version$version.string, "\n")
cat("Platform:", R.version$platform, "\n")
if (requireNamespace("Chromatograms", quietly = TRUE)) {
  cat("Chromatograms Version:",
      as.character(packageVersion("Chromatograms")), "\n")
}
```

## Motivation

The `Chromatograms` package stores chromatographic peaks data as a list of
`data.frame`s — one per chromatogram — rather than a single large matrix.
This analysis provides empirical justification for that design choice.

Chromatographic analysis (peak picking, integration, smoothing, similarity
scoring) operates primarily on vectors of intensity or retention time within
a single chromatogram. The alternative matrix layout would optimize for
cross-chromatogram comparisons at the cost of the far more frequent
single-chromatogram operations.

We benchmark:

1. Columnar access — extracting the intensity vector from a single
   chromatogram (the dominant operation).
2. Row access — extracting a single data point across all variables
   (rare in practice).
3. Scaling behaviour — how the advantage holds as chromatogram length grows.

## Overview: Measurement Scales

All timing measurements in this document use nanoseconds (ns) for reporting,
which is appropriate for per-operation latencies in the microsecond to
nanosecond range. Memory measurements use megabytes (MB) for practical
discussion of object footprints.
```{r}
#| label: load-microbenchmark
library(microbenchmark)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
#| label: inline-helpers
#| include: false
## Helpers for the numbers quoted in the prose. Medians are taken from
## summary(<microbenchmark>) in nanoseconds and rounded to the nearest 100 ns;
## ratios are rounded to one decimal.
med_ns <- function(mb, expr) {
  s <- summary(mb, unit = "ns")
  s$median[as.character(s$expr) == expr]
}
scaling_med <- function(structure, n) {
  scaling_summary$median_ns[scaling_summary$expr == structure &
                              scaling_summary$n_points == n]
}
speedup_df <- function(n) scaling_med("matrix", n) / scaling_med("df", n)
fmt_ns <- function(x) formatC(round(x, -2), format = "d", big.mark = ",")
fmt_x  <- function(x) formatC(x, format = "f", digits = 1, big.mark = ",")
## Median of S4 slot access via `@`, measured in a separate benchmark
## (not run in this document); used only for the appendix comparison.
s4_slot_ns <- 200
```

## 1. Basic Column Access Benchmark

We simulate a single chromatogram of typical length (1 000 data points) stored
as either a matrix, a data.frame, or a data.table, then measure columnar
extraction.

```{r}
#| label: basic-benchmark
set.seed(42)
n_rows <- 1000L

# Matrix representation
mat <- matrix(runif(n_rows * 2), ncol = 2)
colnames(mat) <- c("rtime", "intensity")

# data.frame representation (what Chromatograms uses)
df <- data.frame(rtime = runif(n_rows), intensity = runif(n_rows))

# data.table representation
dt <- as.data.table(df)

mb_basic <- microbenchmark(
  matrix_col_name   = mat[, "intensity"],
  matrix_col_index  = mat[, 2L],
  df_dollar         = df$intensity,
  df_bracket        = df[["intensity"]],
  dt_dollar         = dt$intensity,
  dt_bracket        = dt[["intensity"]],
  times = 10000L,
  unit = "nanoseconds"
)

mb_basic
```

```{r}
#| label: fig-basic-benchmark
#| fig-cap: "Column access latency: matrix vs data.frame vs data.table (1 000 rows)"
autoplot(mb_basic) +
  labs(
    x = "Time (ns)", y = NULL,
    title = "Column Access: Matrix vs data.frame vs data.table"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: basic-headline
## Medians of this run and the storage-layout ratio quoted in the thesis:
## named matrix column extraction relative to data.frame `$` access.
med_basic <- sapply(split(mb_basic$time, mb_basic$expr), median)
knitr::kable(
  data.frame(accessor = names(med_basic), median_ns = med_basic,
             relative_to_df_dollar = med_basic / med_basic[["df_dollar"]]),
  digits = 2, row.names = FALSE,
  caption = "Column access, 1 000 rows: medians of this run (ns)")
cat(sprintf("Storage-layout ratio (matrix_col_name / df_dollar): %.2f\n",
            med_basic[["matrix_col_name"]] / med_basic[["df_dollar"]]))
```

### Interpretation

The results show an advantage for data.frame column access via the $ operator
(median ≈ `r fmt_ns(med_ns(mb_basic, "df_dollar"))` ns) compared to named matrix column extraction
(median ≈ `r fmt_ns(med_ns(mb_basic, "matrix_col_name"))` ns) — a `r fmt_x(med_ns(mb_basic, "matrix_col_name") / med_ns(mb_basic, "df_dollar"))`× speedup. Bracket-based data.frame access
(df[["col"]], median ≈ `r fmt_ns(med_ns(mb_basic, "df_bracket"))` ns) performs similarly to matrix indexing,
confirming that the speedup is specific to the $ operator. Internally, $
bypasses dimension checking and performs a direct list-element lookup on the
underlying data.frame structure, whereas mat[, "col"] must resolve dimension
names, validate indices, and allocate a result vector.

data.table $ access performs comparably to data.frame $ access, since
data.table inherits from data.frame and the $ operator follows the same
list-element lookup path. The bracket accessor dt[["col"]] likewise performs
similarly to df[["col"]]. This confirms that data.table does not provide an
additional columnar access advantage over data.frame for this simple
extraction pattern — both benefit from the same underlying list structure.

Since chromatographic workflows access rtime and intensity columns thousands
of times during peak detection and integration, this constant-factor
advantage over matrices accumulates.

## 2. Row Access Benchmark

For completeness, we also benchmark row access — extracting a single scan across
all variables. This is the operation where matrices traditionally excel, but it
is rarely needed in chromatographic workflows.
```{r}
#| label: row-benchmark
mb_row <- microbenchmark(
  matrix_row = mat[500L, ],
  df_row     = df[500L, ],
  dt_row     = dt[500L, ],
  times = 10000L,
  unit = "nanoseconds"
)

mb_row
```

```{r}
#| label: fig-row-benchmark
#| fig-cap: "Row access latency: matrix vs data.frame vs data.table"
autoplot(mb_row) +
  labs(
    x = "Time (ns)", y = NULL,
    title = "Row Access: Matrix vs data.frame vs data.table"
  ) +
  theme_minimal(base_size = 13)
```

### Interpretation

Matrix row access (median ≈ `r fmt_ns(med_ns(mb_row, "matrix_row"))` ns) is substantially faster than
data.frame row access (median ≈ `r fmt_ns(med_ns(mb_row, "df_row"))` ns) — roughly `r fmt_x(med_ns(mb_row, "df_row") / med_ns(mb_row, "matrix_row"))`× faster. This is
because a matrix stores all values in a single contiguous memory block, so
extracting a row is a simple pointer-offset calculation. In contrast, a
data.frame row slice must extract one element from each column list and
assemble them into a new data.frame. data.table row access
(median ≈ `r fmt_ns(med_ns(mb_row, "dt_row"))` ns) is slower still than data.frame row access, because the
data.table `[` method adds argument-processing overhead on top of the same
column-oriented internal storage.

However, this trade-off is deliberate. Chromatographic analysis pipelines
rarely need cross-variable row access. Operations such as peak detection,
integration, FWHM estimation, and similarity scoring all work on vectors of
intensity or rtime. The only scenario where row access might be relevant is
extracting a single scan point across multiple channels, which is uncommon.
Optimizing for the dominant access pattern (columnar) at the cost of the rare
pattern (row-based) is therefore the correct engineering decision.

## 3. Scaling Behaviour

How does the column-access advantage scale with chromatogram length?

```{r}
#| label: scaling-benchmark
sizes <- c(100, 500, 1000, 5000, 10000, 50000)

scaling_results <- lapply(sizes, function(n) {
  m <- matrix(runif(n * 2), ncol = 2)
  colnames(m) <- c("rtime", "intensity")
  d <- data.frame(rtime = m[, 1], intensity = m[, 2])
  dtb <- as.data.table(d)

  mb <- microbenchmark(
    matrix     = m[, "intensity"],
    df         = d$intensity,
    data.table = dtb$intensity,
    times  = 5000L,
    unit   = "nanoseconds"
  )

  mb |>
    as.data.frame() |>
    mutate(n_points = n)
})

scaling_df <- bind_rows(scaling_results)
```

```{r}
#| label: fig-scaling
#| fig-cap: "Column access time vs chromatogram length"
scaling_summary <- scaling_df |>
  group_by(expr, n_points) |>
  summarise(
    median_ns = median(time),
    q25 = quantile(time, 0.25),
    q75 = quantile(time, 0.75),
    .groups = "drop"
  )

ggplot(scaling_summary, aes(x = n_points, y = median_ns, colour = expr)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = expr), alpha = 0.15,
              colour = NA) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    x = "Chromatogram Length (data points)",
    y = "Median Access Time (ns)",
    colour = "Structure",
    fill = "Structure",
    title = "Column Access Scaling: Matrix vs data.frame vs data.table"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: speedup-table
scaling_summary |>
  select(expr, n_points, median_ns) |>
  pivot_wider(names_from = expr, values_from = median_ns) |>
  mutate(
    speedup_df = matrix / df,
    speedup_dt = matrix / data.table
  ) |>
  knitr::kable(
    digits = 1,
    col.names = c("N points", "Matrix (ns)", "data.frame (ns)",
                  "data.table (ns)", "Speedup df (×)", "Speedup dt (×)"),
    caption = "Column access speedup over matrix"
  )
```

### Interpretation

The scaling results demonstrate that the data.frame column
access time is essentially constant regardless of the number of rows
(≈ `r fmt_ns(min(scaling_summary$median_ns[scaling_summary$expr == "df"]))`–`r fmt_ns(max(scaling_summary$median_ns[scaling_summary$expr == "df"]))` ns across all sizes), while matrix column extraction time grows
linearly with chromatogram length (from ≈ `r fmt_ns(scaling_med("matrix", 100))` ns at 100 points to
≈ `r fmt_ns(scaling_med("matrix", 50000))` ns at 50 000 points).

This is because $ returns a reference to the existing column vector without
copying, whereas mat[, col] must allocate a new vector and copy the selected
column from the contiguous matrix memory block. As chromatogram length
increases, the copy cost grows proportionally.

The practical implication is significant: at 5 000 data points (typical for
UHPLC data), data.frame access is `r fmt_x(speedup_df(5000))`× faster; at 50 000 points (common for
high-resolution full-scan MS data), the speedup reaches `r fmt_x(speedup_df(50000))`×. This means the
data.frame design becomes more advantageous precisely for the high-resolution
data where performance matters most.

## 4. List of data.frames vs data.tables vs Matrices (Realistic Scenario)

In a real Chromatograms object, we store many chromatograms as a list. The
realistic design question is: should each element be a data.frame (with
named columns rtime, intensity), a data.table, or a 2-column matrix?

We compare these three plausible designs for the dominant operation — extracting
the intensity vector of a single chromatogram from the collection.

```{r}
#| label: realistic-benchmark
n_chrom <- 1000L
n_pts   <- 1000L

# List of data.frames (actual Chromatograms design)
list_df <- lapply(seq_len(n_chrom), function(i) {
  data.frame(
    rtime     = sort(runif(n_pts, 0, 600)),
    intensity = abs(rnorm(n_pts, mean = 1000, sd = 500))
  )
})

# List of matrices (plausible alternative)
list_mat <- lapply(list_df, function(d) {
  m <- as.matrix(d)
  colnames(m) <- c("rtime", "intensity")
  m
})

# List of data.tables (plausible alternative)
list_dt <- lapply(list_df, as.data.table)

mb_realistic <- microbenchmark(
  list_df_dollar    = list_df[[500]]$intensity,
  list_df_bracket   = list_df[[500]][["intensity"]],
  list_dt_dollar    = list_dt[[500]]$intensity,
  list_dt_bracket   = list_dt[[500]][["intensity"]],
  list_mat_name     = list_mat[[500]][, "intensity"],
  list_mat_index    = list_mat[[500]][, 2L],
  times = 10000L,
  unit = "nanoseconds"
)

mb_realistic
```

```{r}
#| label: fig-realistic
#| fig-cap: "Accessing intensity of one chromatogram from a collection of 1 000"
autoplot(mb_realistic) +
  labs(
    x = "Time (ns)", y = NULL,
    title = "List of data.frames vs data.tables vs Matrices"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: realistic-headline
med_real <- sapply(split(mb_realistic$time, mb_realistic$expr), median)
knitr::kable(
  data.frame(accessor = names(med_real), median_ns = med_real,
             relative_to_list_df_dollar = med_real / med_real[["list_df_dollar"]]),
  digits = 2, row.names = FALSE,
  caption = "One chromatogram from a list of 1 000: medians of this run (ns)")
```

### Interpretation

Even in a realistic list-based scenario with 1 000 chromatograms, data.frame
$ access (median ≈ `r fmt_ns(med_ns(mb_realistic, "list_df_dollar"))` ns) remains the fastest approach — `r fmt_x(med_ns(mb_realistic, "list_mat_index") / med_ns(mb_realistic, "list_df_dollar"))`× faster
than the best matrix accessor (mat[, 2] by integer index, median ≈ `r fmt_ns(med_ns(mb_realistic, "list_mat_index"))` ns)
and `r fmt_x(med_ns(mb_realistic, "list_mat_name") / med_ns(mb_realistic, "list_df_dollar"))`× faster than named matrix access (mat[, "intensity"],
median ≈ `r fmt_ns(med_ns(mb_realistic, "list_mat_name"))` ns). data.table $ access performs comparably to data.frame $
access, confirming that the list-element lookup mechanism is shared.
Bracket-based access for both data.frame and data.table performs comparably
to matrix indexing, confirming that the advantage is specific to the $
operator.

This benchmark is the most directly relevant to the Chromatograms package
design, since it mirrors the actual storage layout: a list of per-chromatogram
containers. The results confirm that the list-of-data.frames architecture
provides consistently faster columnar access than a list-of-matrices, while
data.table offers no additional advantage for this access pattern. Since
data.table adds dependency weight without a performance benefit for
per-chromatogram column extraction, the data.frame choice is the most
parsimonious.

## 5. Memory Overhead Comparison

The list-of-data.frames representation also has implications for memory. Notably,
it supports variable-length chromatograms without padding, which is critical
for real data where chromatograms span different RT ranges. We compare the
memory footprint of list-of-data.frames vs list-of-matrices for both
fixed-length and variable-length chromatograms.

```{r}
#| label: memory-comparison
# Variable-length chromatograms (realistic)
set.seed(42)
var_list_df <- lapply(seq_len(n_chrom), function(i) {
  n <- sample(200:2000, 1)
  data.frame(
    rtime     = sort(runif(n, 0, 600)),
    intensity = abs(rnorm(n, 1000, 500))
  )
})

var_list_mat <- lapply(var_list_df, function(d) {
  m <- as.matrix(d)
  colnames(m) <- c("rtime", "intensity")
  m
})

var_list_dt <- lapply(var_list_df, as.data.table)

# List of data.tables (fixed-length, from earlier)
list_dt_fixed <- lapply(list_df, as.data.table)

get_size_mb <- function(x) {
  as.numeric(object.size(x)) / 1024^2
}

memory_df <- data.frame(
  Scenario = c("Fixed-length (1 000 × 1 000 pts)",
               "Variable-length (1 000 × 200–2 000 pts)"),
  `data.frame (MB)` = round(c(get_size_mb(list_df),
                               get_size_mb(var_list_df)), 1),
  `data.table (MB)` = round(c(get_size_mb(list_dt_fixed),
                               get_size_mb(var_list_dt)), 1),
  `matrix (MB)` = round(c(get_size_mb(list_mat),
                           get_size_mb(var_list_mat)), 1),
  check.names = FALSE
)

knitr::kable(
  memory_df,
  caption = "Memory footprint comparison across storage representations"
)
```

### Interpretation

The memory comparison shows that data.frame and data.table overhead relative
to matrices is negligible in practice — all three list-based representations
have comparable footprints. The small overhead arises from per-element
metadata (column names, row names, class attributes) stored once per
chromatogram — a constant cost dwarfed by the actual data vectors.
data.table's additional internal attributes (e.g., `.internal.selfref`) add
a marginal cost that is not meaningfully different at this scale.

Crucially, none of the list-based representations require padding. A single
large matrix storing all chromatograms would need every row to have the same
number of columns, forcing shorter chromatograms to be padded with NA values.
For a dataset where chromatogram lengths vary from 200 to 2 000 points,
padding would inflate memory usage substantially. The list-of-data.frames
design avoids this waste entirely while paying virtually no overhead compared
to list-of-matrices or list-of-data.tables.

## Conclusion

The benchmarks confirm that the list-of-data.frames design is well suited for
the Chromatograms package. The `$` accessor provides `r fmt_x(speedup_df(100))`–`r fmt_x(speedup_df(50000))`× faster column
extraction than matrices (scaling with chromatogram length), while data.table
offers no additional speed advantage since both share the same list-element
lookup. Memory overhead across all three representations is negligible, and
only the list-based designs handle variable-length chromatograms without
wasteful padding. The sole trade-off — slower row access compared to
matrices — is irrelevant in practice, as chromatographic workflows operate
on per-column vectors. Choosing data.frame over data.table avoids an external
dependency with no performance cost.

## Appendix: Why Not S4 Slots?

A natural alternative to data.frames for per-chromatogram storage is an S4
class with dedicated slots — e.g., `rtime` and `intensity` as separate
numeric vectors. Micro-benchmarks (available in a separate branch) show that
S4 slot access via `@` is indeed the fastest read accessor (~`r fmt_ns(s4_slot_ns)` ns median),
roughly `r fmt_x(med_ns(mb_basic, "df_dollar") / s4_slot_ns)`× faster than `data.frame` `$` access. However, several structural
constraints make S4 unsuitable for the `peaksData` representation.

**Fixed schema.** An S4 class must declare its slots at definition time.
The `peaksData` contract in `Chromatograms` supports arbitrary peaks
variables, not just `rtime` and `intensity`, but also `mz` (for 3D
chromatograms), annotation columns, or any user-defined variable. Adding a
new variable to an S4 class requires redefining the class itself; adding a
column to a data.frame is trivial. Moreover, different chromatograms within
the same collection may carry different sets of peaks variables, something
data.frames handle natively but S4 cannot.

**Access method overhead.** While raw `@` access is fast, S4 objects in a
package API are never accessed via `@` directly — they require formal
accessor methods (`intensity()`, `rtime()`, etc.). Each accessor call
involves S4 method dispatch, which adds overhead per invocation. In a
pipeline that repeatedly accesses intensity and rtime across thousands of
chromatograms, this repeated dispatch erodes the raw slot-access advantage.
data.frame `$` access requires no method dispatch: it is a direct
list-element lookup.

**Paired access pattern.** Chromatographic operations typically consume
`rtime` and `intensity` together (e.g., peak detection, integration,
plotting). A data.frame naturally groups these as columns of the same
object, accessible in a single `peaksData()[[i]]` call. With S4 slots, each
variable requires a separate accessor, doubling the dispatch overhead for the
most common access pattern.

**Nested S4 scalability.** A `Chromatograms` object stores hundreds to
thousands of chromatograms. Storing these as a `list` of S4 objects
introduces per-element S4 overhead (class metadata, validity infrastructure)
that compounds across the collection. R's garbage collector and serialization
routines also handle large lists of S4 objects less efficiently than lists of
data.frames, which are simple named lists internally.

In summary, S4 slots offer faster raw read access but cannot match the
flexibility, interoperability, and ecosystem compatibility that data.frames
provide. The ~`r fmt_ns(med_ns(mb_basic, "df_dollar") - s4_slot_ns)` ns per-access difference is irrelevant at pipeline scale.

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
